---
title: Testing the Central Limit Theorem with the R Exp distribution (A Simulation
  Exercise)
author: "Juan Jose"
date: "Monday, March 16, 2015"
output: word_document
---



## Overview
This analysis seeks to compare the exponential distribution in R, rexp, with the Central Limit Theorem. To do so, we will compare the sample means of 40 exponentials a total of 1000 times.

First load all the necessary libraries
```{r}
library(ggplot2)
```

## Simulations
For the purpose of this study, we will define the rate parameter of the R rexp function as 0.2
```{r}
lambda = 0.2
#this is a place holder for the means we will collect during the simulations.
all.the.means = NULL
#num.sims is the number of simulations we will run, in this case 1000
num.sims = 1000

#set a seed to make it reproducible
set.seed(500)

for(i in 1:num.sims) all.the.means = c(all.the.means, mean(rexp(40,lambda)))

#store the simulation results in a data frame
df = data.frame(x=1:num.sims, y=all.the.means)

```

## Analyzing The Mean (Answer to Question 1)

We know the population mean of the exponential distribution is 1/lambda. In our case, lambda = 0.2 so the theoretical mean is 5. 

### Theoretical Mean
```{r}
pop.mean = 1/lambda
pop.mean
```

### Mean of all the Simulations
```{r}
sample.mean = mean(all.the.means)
sample.mean
```

Our sample mean is close to the theoretical mean.

## Analyzing The Variance (Answer to Question 2)


Since our lambda is 0.2, we know our theoretical sample variance is 

### Sample Variance
```{r}
theoretical.var = ((1/lambda)/sqrt(40))^2
theoretical.var
```

### Variance of the Simulations
```{r}
sample.var = var(all.the.means)
sample.var
```

Our sample variance to the theoretical variance.

## Show that the distribution is approximately normal.


From the histogram, we can see that the shape resembles that of a normal distribution with a mean of (1/lambda)
and standard deviation of (1/lambda)/sqrt(40)), where lambda = 0.2.
```{r}
#histgoram of the means
g <- ggplot(df, aes(x=y)) + geom_histogram(aes(y=..density..), binwidth=0.1, fill="red", color="black") + labs(title="Plot of the Simulations")
g = g + stat_function(fun=dnorm, arg=list(mean=(1/lambda), sd=(1/lambda)/sqrt(40)), size=2) 
print(g)

```
